\Title

### 2013-05-29 synmut.tex
More discussion on RNA stuff + citations

A functional significance of the insulating RNA structure stems between the
hypervariable loops has also been proposed previously
\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
structures exist in different parts of the HIV-1 genome. Since there are
of course many ways to build an RNA stem in a particular location, we do
not necessarily expect a strong signal of conservation in cross-sectional data. 
Our analysis, however, is able to
quantify the fitness effect of RNA structure within single infections and
demonstrates how selection at synonymous sites can alter genetic diversity and
dynamics. %FIXME insert citations and more discussion

@@

Alternatively, the lack of fixation could be due to time-dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect in computer models. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV-1
infected individuals typically neutralizes the virus that dominated the
population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large effective population size that explores all necessary point
mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012}.

>>>>

A functional significance of the insulating RNA structure stems between the
hypervariable loops has also been proposed previously
\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
structures exist in different parts of the HIV-1 genome. Since there are of
course many ways to build an RNA hairpin at a particular location, we do not
necessarily expect a strong signal of conservation in cross-sectional data.
Consistently, in a characterization of the SIV RNA structures via the SHAPE
assay, it was recently shown that only the general pattern is shared with HIV,
but the single base pairs are often discordant between the two viruses
\citep{pollom_comparison_2013}. In fact, for each pairing-disrupting mutation in
our dataset, we have looked for compensatory mutations at the partner site, but
observed no effect. For these reasons, we do not believe that exact RNA pairings
are evolutionary relevant, rather only the general architecture of different
genomic regions (e.g. the V loops). Despite this flexibility in pairing, our
analysis is able to quantify the fitness effect of RNA structure within single
infections and demonstrates how selection at synonymous sites can alter genetic
diversity and dynamics.

In principle, the fitness effect of RNA structures in terms of replication
capacity could be tested culturing \textit{in vitro} viruses with engineered
modifications. This has been attempted recently, but no major fitness effect was
measured \citep{knoepfel_role_2013}. There is no contradiction with our results,
because we infer a selection coefficient for each mutation of the order of
$-0.002$, which is only visible after hundreds of generations. Shorter
experimental tests cannot detect such a subtle effect; our long-term
longitudinal analysis is more sensitive in this respect.

@@

Alternatively, the lack of fixation could be due to time-dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect in computer models. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV-1
infected individuals typically neutralizes the virus that dominated the
population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large effective population size that explores all necessary point
mutations rapidly. A similar point has been made recently by several authors
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012, pennings_loss_2013}.
